A Linear Solution of the Linearity Challenge

### 2025-12-24 paper/Type.lagda.tex
revised congruence
cleanup
revised types

The representation of types is standard, with one constructor for each of the
forms described in \Cref{sec:calculus}. We start by defining \emph{pre-types} as
an indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
type variables and we use de Bruijn indices for denoting type variables. In this
way, we make sure that pre-types are well-scoped.

@@

This property is so important in the rest of the formalization that it is worth
defining an implicit rewriting rule so that Agda can autonomously apply it
whenever possible. This is achieved by means of the following pragma
directive.\footnote{The directive is effective provided that the option
\texttt{--rewriting} is enabled, either globally when invoking Agda or in the
module's preamble.}

@@

Next we define the function \AgdaFunction{subst} that simultaneously substitutes
the type variables of a (pre)type with other (pre)types. Even though in practice
we will always substitute one variable at a time, it is technically easier to
substitutes \emph{all} variables of a (pre)type. The definition of
\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
variables and \emph{lifting} substitutions across quantifiers. All such
definitions are straightforward adaptations of those described by
\cite{KokkeWadler}.

@@

\begin{code}
[_/_] : ∀{n} → PreType n → Fin (suc n) → PreType n
[ A / zero  ]  = A
[ A / suc k ]  = var k
\end{code}

@@

It is worth discussing one case in the proof of \AgdaFunction{dual-subst}:
\begin{code}
dual-subst {_} {_} {σ} {rav x} = refl
\end{code}

Here we are supposed to prove
$\AgdaFunction{subst}~\sigma~(\AgdaFunction{dual}~(\AgdaInductiveConstructor{rav}~x))
  \equiv
  \AgdaFunction{dual}~(\AgdaFunction{subst}~\sigma~(\AgdaInductiveConstructor{rav}~x))$
  which is definitionally the same of $\sigma~x \equiv
  \AgdaFunction{dual}~(\AgdaFunction{dual}~(\sigma~x))$.
%
We could easily prove this equivalence by invoking \AgdaFunction{dual-inv}, but
thanks to the added rewriting rule a use of \AgdaInductiveConstructor{refl}
suffices. In this case the saved effort is negligible, but in later results,
where it is necessary to use \AgdaFunction{dual-inv} for rewriting part of the
\emph{index} of some type families, having an implicit rewriting rule allows us
to avoid writing some quite obscure Agda code.

@@

Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
formalization that follows. Therefore, we add it to the set of implicit
rewriting rules used by Agda so that we do not have to think about it again.

>>>>

The representation of types is standard, with one constructor for each of the
forms described in \Cref{sec:types}. We start by defining \emph{pre-types} as an
indexed data type $\AgdaDatatype{PreType}~n$ where $n$ is the number of bound
type variables and we use de Bruijn indices for denoting type variables. In this
way, we make sure that pre-types are well-scoped.

@@

This property is so important in the rest of the formalization that it is worth
defining an implicit rewriting rule so that Agda can autonomously apply it
whenever possible. This is achieved by means of the following pragma
directive.\footnote{The directive is effective provided that the option
\texttt{{-}{-}rewriting} is enabled, either globally when invoking Agda or in
the module's preamble.}

@@

Next we define the function \AgdaFunction{subst} that simultaneously substitutes
the type variables of a (pre)type with other (pre)types. Even though in practice
we will always substitute one variable at a time, it is technically easier to
substitutes \emph{all} variables of a (pre)type. The definition of
\AgdaFunction{subst} relies on some auxiliary functions for \emph{renaming} type
variables and \emph{lifting} substitutions across quantifiers. All such
definitions are straightforward adaptations of those described by
\citet{KokkeSiekWadler20}.

@@

\begin{code}
[_/] : ∀{n} → PreType n → Fin (suc n) → PreType n
[ A /] zero     = A
[ A /] (suc k)  = var k
\end{code}

@@

It is worth looking at one case in the proof of \AgdaFunction{dual-subst},
namely when the type is a dualized type variable:

\begin{code}
dual-subst {_} {_} {σ} {rav x} = refl
\end{code}

Here we are supposed to prove
$\AgdaFunction{subst}~\sigma~(\AgdaFunction{dual}~(\AgdaInductiveConstructor{rav}~x))
  \equiv
  \AgdaFunction{dual}~(\AgdaFunction{subst}~\sigma~(\AgdaInductiveConstructor{rav}~x))$
  which is definitionally the same of $\sigma~x \equiv
  \AgdaFunction{dual}~(\AgdaFunction{dual}~(\sigma~x))$.
%
We could easily prove this equivalence by invoking \AgdaFunction{dual-inv}, but
thanks to the rewriting rule that we have added just above a use of
\AgdaInductiveConstructor{refl} suffices. In this case the saved effort is
negligible, but in later results, where it is necessary to use
\AgdaFunction{dual-inv} for rewriting part of the \emph{index} of some type
families, having an implicit rewriting rule allows us to avoid writing some
quite obscure Agda code.

@@

Just like \AgdaFunction{dual-inv}, \AgdaFunction{dual-subst} too is key in the
formalization that follows. Therefore, we add it to the set of implicit
rewriting rules used by Agda so that we do not have to think about this property
again.

### 2025-12-24 paper/agda.tex
included reduction
drafted deadlock freedom
drafted type safety
(re)started paper
revised types
restroed contexts and permutations
new notation for splitting
restored DF
restored type safety
restructuring presentation with new code
revised congruence
restored reduction
restored examples

\input{Type}
\input{Context}
\input{Permutations}
\input{Process}
\input{Semantics}

>>>>

In this section we describe the formalization of \LCP in Agda. We present in
detail only the key parts of the code, including the representation of types,
processes, the definition of the operational semantics and the statement of the
main results. The interested reader may refer to the public repository for the
complete source code.

\input{Type}
\input{Context}
\input{Permutations}
\input{Process}
\input{Congruence}
\input{Reduction}
\input{DeadlockFreedom}
\input{Safety}
\input{Examples}

### 2025-12-24 paper/calculus.tex
revised types

As we have anticipated in \Cref{sec:introduction}, \LCP is a linear version of
\CP~\citep{Wadler14,GayVasconcelos25} such that every linear channel is meant to
be used for \emph{exactly one} communication. This differs from \CP and other
session calculi where session endpoints can be used repeatedly, albeit
sequentially. A session consisting of several messages exchanged in sequence can
be encoded in \LCP using linear channels by pairing each message with a
\emph{fresh continuation channel} on which the rest of the session takes place.
%
This way of encoding sessions using explicit continuations has been first
described by \citet[extended version]{Kobayashi02b} and later studied by
\citet{DardhaGiachinoSangiorgi17}.

In this section we present \LCP, starting from its types (\Cref{sec:types}),
which coincide with classical linear logic propositions, its syntax
(\Cref{sec:syntax}), the reduction semantics (\Cref{sec:semantics}), the typing
rules (\Cref{sec:typing-rules}) and the formulation of the properties ensured by
the type system (\Cref{sec:properties}).

>>>>

In this section we give a cursory presentation of \LCP.
%
As we have anticipated in \Cref{sec:introduction}, \LCP is very closely related
to \CP~\citep{Wadler14,GayVasconcelos25} except that \LCP features linear
channels instead of sessions. A session consisting of several messages exchanged
in sequence can be encoded in \LCP using linear channels by pairing each message
with a \emph{fresh continuation channel} on which the rest of the session takes
place.
%
This way of encoding sessions using explicit continuations has been first
described by \citet[extended version]{Kobayashi02b} and later studied by
\citet{DardhaGiachinoSangiorgi17}.

We present \LCP starting from its types (\Cref{sec:types}), which coincide with
classical linear logic propositions, then moving on to the syntax of processes
(\Cref{sec:syntax}), their reduction semantics (\Cref{sec:semantics}), the
typing rules (\Cref{sec:typing-rules}) and the formulation of the properties
ensured by the type system (\Cref{sec:properties}).
